---
title: "Exploring ECC counter data"
author: "Diarmuid Lloyd"
date: "Last run: `r Sys.Date() `"
output: 
  html_document:
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dbplyr)
library(DBI)

```

# Introduction

Loading counter data for locations John flagged as of interest:

**On Road - SfP**

* 031 Dundee Street
* 035 A90 Deans Bridge
* 048 Crewe Rd South
* 038 Bruntsfield Pl (Southbound)

**On Road - Non SfP**

* 052 A8 Wester Coates

**Off Road - Non SfP**

* 015 Middle Meadow Walk

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = "data/ecc-counter-database-output.sqlite")

df <- tbl(con, "station_counts_hour") %>% 
  filter(station_num %in% c("31","35","48","38","52","15")) %>% collect()

dbDisconnect(con)

```

# Intial plots and exploring data

```{r}

df <- df %>% mutate(date = lubridate::as_datetime(date))

df %>% count(station_name)

df %>% count(flag_text)

df <- df %>% 
  group_by(station_num) %>% 
  mutate(week_starting = lubridate::floor_date(date, unit = "week") %>% as.Date(),
         year = lubridate::year(date),
         month = lubridate::month(date))

df_weekly <- df %>% 
  group_by(station_name,week_starting) %>% 
  filter(is.na(flag_text)) %>% 
  summarise(volume = sum(volume))

```

Plotting raw volume by hour

```{r}

ggplot(df) +
  geom_line(aes(x = date, y = volume, group = station_num,colour = station_name)) +
  facet_wrap(~station_name) +
  theme(legend.position = "none")

```

Plotting weekly total trips, also filtering out all flagged "bad data". Seems to knock out Pedal on Parliament on MMW? Regardless, huge gaps in the data.

```{r}

ggplot(df_weekly) +
  geom_col(aes(week_starting, volume, fill = station_name)) + 
  facet_wrap(~station_name) +
    theme(legend.position = "none")

```



