---
title: "Exploring ECC counter data"
author: "Diarmuid Lloyd"
date: "Last run: `r Sys.Date() `"
output: 
  html_document:
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dbplyr)
library(DBI)

```

# Introduction

Loading counter data for locations John flagged as of interest:

**On Road - SfP**

* 031 Dundee Street
* 035 A90 Deans Bridge
* 048 Crewe Rd South
* 038 Bruntsfield Pl (Southbound)

**On Road - Non SfP**

* 052 A8 Wester Coates

**Off Road - Non SfP**

* 015 Middle Meadow Walk

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = "data/ecc-counter-database-output.sqlite")

df <- tbl(con, "station_counts_hour") %>% 
  filter(station_num %in% c("31","35","48","38","52","15")) %>% collect()

dbDisconnect(con)

```

# Intial plots and exploring data

```{r}

df <- df %>% mutate(date = lubridate::as_datetime(date))

df %>% count(station_name)

df %>% count(flag_text)

df <- df %>% 
  group_by(station_num) %>% 
  mutate(week_starting = lubridate::floor_date(date, unit = "week") %>% as.Date(),
         year = lubridate::year(date),
         month_starting = lubridate::floor_date(date, unit = "month") %>% as.Date())

df_weekly <- df %>% 
  group_by(station_name,week_starting) %>% 
  filter(is.na(flag_text)) %>% 
  summarise(volume = sum(volume))


df_monthly <- df %>% 
  group_by(station_name,month_starting) %>% 
  filter(is.na(flag_text)) %>% 
  summarise(volume = sum(volume))

```

Plotting raw volume by hour

```{r}

ggplot(df) +
  geom_line(aes(x = date, y = volume, group = station_num,colour = station_name)) +
  facet_wrap(~station_name) +
  theme(legend.position = "none")

```

## Filtered out bad data

Plotting weekly total trips, also filtering out all flagged "bad data". Seems to knock out Pedal on Parliament on MMW (guessed this is what the peaks are)? Regardless, huge gaps in the data.

```{r}

weekly_counts_rough <- ggplot(df_weekly) +
  geom_col(aes(week_starting, volume, fill = station_name)) + 
  facet_wrap(~station_name) +
    theme(legend.position = "none")


weekly_counts_rough
png("output/rough_weekly_counts.png",width = 30, height = 20, units = "cm", res = 300 )
weekly_counts_rough
dev.off()
```

Plotting monthly total trips, again bad data filtered out, but also clipping to 2019 onward:

```{r}

monthly_counts_rough <- ggplot(df_monthly) +
  geom_col(aes(month_starting, volume, fill = station_name)) + 
  facet_wrap(~station_name) +
    theme(legend.position = "none") +
  scale_x_date(limits = c(as.Date("2019-01-01"),as.Date("2022-01-01")),
               name = "Monthly counts")

monthly_counts_rough
png("output/rough_monthly_counts.png", width = 30, height = 20, units = "cm", res = 300 )
monthly_counts_rough
dev.off()


```





